\Large \bf 6.824 Final Project

### 2014-05-11 writeup/paper.tex
changed debugging
Persistence section draft (may be too much info)

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf 6.824 Final Project}
%for single author (just remove % characters)
\author{
{\rm Colleen Josephson}\\
cjoseph@mit.edu
\and
{\rm Joseph DelPreto}\\
delpreto@mit.ed
\and
{\rm Pranjal Vachaspati}\\
pranjal@mit.edu
\and
{\rm Steven Valdez}\\
dvorak42@mit.edu
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
} % end author

@@

\section{Introduction}
This document is the writeup of our 6.824 final project. We did the
default project, and built a persistent, fault-tolerant
high-performance key/value store. For persistence, we used the go
wrapper for the high-performance C++ leveldb key/value libray, and
extended our code to recover from data loss upon server restart. In
order to improve the performance of our system, we implemented two
paxos optimizations. We used a multi-paxos like design to avoid the
dueling leaders problem, and also reduce the number of RPCs for paxos
agreeement by having the leader send out prepare messages ahead of
time. To measure performance, we extended our existing tests, and
added new ones. We also modified the RPC system to work on a real
network and deployed our project on an AWS cluster to get realistic
throughput and latency measures, as well as information about the
speed of disk recovery.

\section{Design} \label{sec:design}
Our starting point was lab 4. Before begining, we ran battery tests to
figure out which combination of our prior labs created the fastest and
most reliable starting point.

@@

\subsection{Persistent Storage}

\subsubsection{Recovery from failures}

>>>>

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf 6.824 Final Project}
%for single author (just remove % characters)
\author{
{\rm Colleen Josephson}\\
cjoseph@mit.edu
\and
{\rm Joseph DelPreto}\\
delpreto@mit.edu
\and
{\rm Pranjal Vachaspati}\\
pranjal@mit.edu
\and
{\rm Steven Valdez}\\
dvorak42@mit.edu
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
} % end author

@@

\section{Introduction}
The presented project implements a persistent, fault-tolerant, high-performance key/value store. In order to achieve these goals, the system implements at-most-once semantics, relies upon Paxos for reaching consenus, shards the database across multiple replica groups, persistently stores values to disk, and implements protocols for automatically recovering and updating disk contents upon server reboot.  In order to improve performance, the system features two paxos optimizations: a protocol similar to multi-paxos to avoid the dueling leaders problem, and the ability for leaders to send out prepare messages in advance to reduce the required number of RPCs for paxos agreeement.

Extensive testing was done to evaluate the system's persistence, recovery, and performance by extending existing tests and creating new ones.  After modifying the local RPC system to work on a real network, the system was deployed on an Amazon Web Services (AWS) cluster to obtain realistic measurements of throughput, latency, and disk recovery speed.

\section{Design} \label{sec:design}
The design is based upon the code developed throughout the semester for lab 4.  There are therefore three basic components running on each server: Paxos processes to reach consenus for instances, a shard master to determine which groups are responsible for various shards of the database, and a shard key/value storage process to execute client Put/Get requests.  The Paxos instances are used to create a consistent log of operations for both shard reconfiguration and client operations.

Battery tests were run to determine which combination of our prior labs created the fastest and most reliable starting point.  Once a working codebase was compiled, modifications were made to address three  major areas: Paxos optimizations, persistence (with disk recovery), and testing over a real network on AWS.

@@

\subsection{Persistent Storage}
In order to tolerate servers restarting and to handle databases that cannot fit in memory, all three components of the system store state and data persistently on disk.  To interface with the disk, a Go wrapper for the high-performance C++ key/value libray LevelDB was utilized.  An outline of what each system component stores on disk is presented below.
\begin{itemize}
\item The Paxos processes write the highest accepted value, the highest accepted proposal number, and the highest acknowledged prepare request to disk for each instance.  This ensures that the protocol will operate correctly even if a server restarts (for example, a decided value can never be changed).  In addition to this state, the Paxos processes persistently store the done responses received from peers and the number of the maximum known instance (see the recovery section below).

\item The shard master writes every configuration to disk, ensuring that will be able to respond to historical queries even after restarting.  In addition, it stores the highest executed Paxos log entry and the number of its highest known configuration.

\item The shard key/value service writes its database to disk by writing each value as each Put request is executed.  In addition, it must preserve at-most-once semantics and consistency across reboots, so it persistently stores which operations have been seen as well as the most recent response sent to each client.  It also writes the highest executed Paxos log entry and the current configuration number.
\end{itemize}

In all cases, the specified data is written to disk before returning to the caller, so any response that a caller receives from a peer will not be forgotten by the responder even if it restarts.  It may be the case that the responder crashes before sending the response, but this is acceptable since the system offers at-most-once semantics (so the caller must be prepared to retry and the responder is prepared to filter duplicates).

\subsubsection{Recovery from Failures}
In order to tolerate servers restarting with or without their disk contents, the system includes protocols for recovering and catching up on startup.  These can be summarized by considering what happens when a server starts:

\begin{itemize}
\item The Paxos process will load any saved done responses and maximum instance number from its disk.   It will then contact peers to request their done responses and maximum instance number, determining which peer is the most updated out of any peers that respond.  From these values, both the ones read from disk and the ones received, it can determine which instances it should know about but does not; it then requests these instances from its peers.
\item The shard master process will load the highest known configuration and highest executed Paxos log instance from disk if it exists.  It will then request the same information from its peers, determining who among the responders is the most updated.  From this information it can determine if there are any missed configurations, and if so request them from a peer.  
\item The shard key/value service will load its last known configuration and the highest executed Paxos log instance from disk if it exists.  It will then request the same information from peers, determining the most updated state that it should have.  Finally, it can request the actual database key/value entries from a peer (see below for details of how this transfer is performed) along with any client responses and lists of seen operations.
\end{itemize}
These protocols allow any number of servers restarting with their disk contents to catch themselves up from any updated peer.  They also allow any number of servers restarting without their disk contents to recover the needed data from any peer that has the data.

\subsubsection{Memory and Performance Considerations}
There are a number of design choices which were made in order to address both memory management and overall performance.

Firstly, the service must be able to handle database content which is too large to fit in volatile memory.  Towards this end, values are written to disk immediately as described above so it does not need to be kept in memory.  More importantly, however, servers that need to transfer shards (either because the shard master decided a shard needs to move or because a server is recovering) must be able to do so in multiple RPCs since memory is too small to create a reply with the entire database.  Therefore, a shard transfer occurrs with the following protocol:
\begin{itemize}
\item The requester sends a request to a peer, including a request number, and the current configuration number.  If doing a shard transfer, the shard number is also included.
\item If the receiver is currently at the appropriate configuration, it accepts the request.  It will then begin populating a reply with client response data, a list of seen operations, and the desired key/value data.  While doing this, it will periodically check memory usage; if it exceeds a configurable threshold, it will reply with whatever it currently has and mark the reply as incomplete.
\item The requester copies the data from the reply to its disk.  If the reply was marked as incomplete, it sends another request to the same peer with an incremented request number.
\item This continues until the reply is marked as complete, at which point the requester launches a goroutine to make sure the sender receives an acknowledgement of the receipt of the final data message.
\item If at any point the requester decides that the contacted peer is dead, it launches an acknowledgement goroutine as previously described for that peer (to terminate the transfer in case the peer wakes up or is not dead) and moves on to try a different peer, resetting the request number to 0.
\end{itemize}
Both the requester and the sender ignore client requests until the transfer is complete - this ensures that the requester will receive a consistent snapshot of the data (and is the reason the requester launches acknowledgement goroutines since otherwise a peer erroneously declared dead would be halted forever).

One aspect which addresses performance is that progress can still be made while recovery is taking place as long as a majority of servers are not recovering.  For example, if a replica group has three servers and a shard transfer is ocurring between two of them (perhaps because one just restarted), a client can still have its requests processed if it contacts the third peer.  This is because the Paxos service runs independently of the shard key/value service; although the latter will ignore client requests on the transferring servers, their Paxos processes can still respond and make progress.  Since the Paxos instances are stored persistently on disk, the two occupied servers can catch themselves up from the log when they complete the transfer.

In addition, effort was made to utilize memory in addition to the disk in order to increase performance.  The LevelDB database has the option to use a least-recently-used (LRU) cache to store recently used entries in memory.  In addition, a configuration variable was added to the service to allow values to be written to memory as well as disk until the memory is full.  These two options, combined with the native cache of the operating system, allow many needed values to be read quickly from memory rather than always consulting the disk.
